Visualizing Multi-way Contingency Tables

### 2005-11-18 trunk/vcd/paper/hsc.Rnw
final fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@501 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% Hyperref
\hypersetup{%
  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed chroma and luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Fig.~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that this varies chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminace better 
contrasts can be achieved.

>>>>

%% Hyperref
\hypersetup{%
  pdftitle = {Visualizing Multi-way Contingency Tables},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
  pdfauthor = {David Meyer, Achim Zeileis, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed chroma and luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Fig.~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminace better 
contrasts can be achieved.
